---
title: Statistical analyses
author: "Matthias Urban"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document: default
---
# Overview

This document describes the procedures for fitting, running, and assessing the Bayesian Mixed Effects Logistic Regression Models.

# Required packages

```{r message=FALSE}
library(tidyverse)
require(brms)
```

# Data preparation

Load in datasets
```{r}
languagesonlandmasswithmountains <- read.csv("../Data/languagesonlandmasswithmountains.csv", header = T, encoding = "UTF-8")
languagesonlandmasswithhighmountains <- read.csv("../Data/languagesonlandmasswithhighmountains.csv", header = T, encoding = "UTF-8")
```
Treat Macroareas as factors
```{r}
languagesonlandmasswithmountains$Macroarea <- as.factor(languagesonlandmasswithmountains$Macroarea)
languagesonlandmasswithhighmountains$Macroarea <- as.factor(languagesonlandmasswithhighmountains$Macroarea)
```

Log 10 transform of predictors after adding a constant to reduce skew
```{r, cache=TRUE}
languagesonlandmasswithmountains <- mutate(languagesonlandmasswithmountains, Family_ID = ifelse(Family_ID == "", Name, Family_ID))
languagesonlandmasswithmountains <- mutate(languagesonlandmasswithmountains, Distancetosealog10 = log10(Distancetosea + 0.001))
languagesonlandmasswithmountains <- mutate(languagesonlandmasswithmountains, Distancetomountainlog10 = log10(Distancetomountain + 0.001))

languagesonlandmasswithhighmountains <- mutate(languagesonlandmasswithhighmountains, Family_ID = ifelse(Family_ID == "", Name, Family_ID))
languagesonlandmasswithhighmountains <- mutate(languagesonlandmasswithhighmountains, Distancetosealog10 = log10(Distancetosea + 0.001))
languagesonlandmasswithhighmountains <- mutate(languagesonlandmasswithhighmountains, Distancetohighmountainlog10 = log10(Distancetohighmountain + 0.001))
```

Set area for the five New Zealand languages in the dataset (the only ones remaining for the Papunesia area) to NA 
```{r}
languagesonlandmasswithhighmountains[which(languagesonlandmasswithhighmountains$Macroarea == "Papunesia"), "Macroarea"] <- NA
languagesonlandmasswithhighmountainsmodeling <- drop_na(languagesonlandmasswithhighmountains, Macroarea)
```

Set priors
```{r, cache=TRUE}
priorshighmountains <- set_prior("normal(0, 2)", class = "b", coef = c("Distancetosealog10", "Distancetohighmountainlog10"))
priorsmountains <- set_prior("normal(0, 2)", class = "b", coef = c("Distancetosealog10", "Distancetomountainlog10"))
```

# Main analysis
```{r, cache=TRUE, warning=FALSE, message=FALSE, results='hide'}
isolatesnarrowhighmountainslog10model <- brm(IsolateNarrow ~ Distancetosealog10 * Distancetohighmountainlog10 + (1 + (Distancetosealog10 * Distancetohighmountainlog10) | Macroarea), family = "bernoulli", data = languagesonlandmasswithhighmountainsmodeling, warmup = 2000, iter = 3000, chains = 4, prior = priorshighmountains, seed = 68, control = list(adapt_delta = 0.999999, max_treedepth = 20))
```

Model assessment and checks

check Rhat and ESS values
```{r}
summary(isolatesnarrowhighmountainslog10model)
```
Inspect chains
```{r}
plot(isolatesnarrowhighmountainslog10model)
```
Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(isolatesnarrowhighmountainslog10model)
pp_check(isolatesnarrowhighmountainslog10model, type = "error_binned")
```
Assess predictive accuracy
```{r, cache=TRUE}
isolatesnarrowhighmountainslog10model_pred <- predict(isolatesnarrowhighmountainslog10model, type = "response")[, "Estimate"]
isolatesnarrowhighmountainslog10model_pred <- as.numeric(isolatesnarrowhighmountainslog10model_pred > mean(languagesonlandmasswithhighmountainsmodeling$IsolateNarrow))
```
```{r, cache=TRUE}
(classtab_isolatesnarrowhighmountainslog10model <- table(predicted = isolatesnarrowhighmountainslog10model_pred, observed = languagesonlandmasswithhighmountainsmodeling$IsolateNarrow))
(acc_isolatesnarrowhighmountainslog10model <- sum(diag(classtab_isolatesnarrowhighmountainslog10model)) / sum(classtab_isolatesnarrowhighmountainslog10model))
```

# Ancillary analyses

## Bernoulli model for isolates in a wide sense and highmountainareas with log10 transformed predictors and random effect for macroarea
```{r, cache=TRUE, warning=FALSE, message=FALSE, results='hide'}
isolateswidehighmountainslog10model <- brm(IsolateWide ~ Distancetosealog10 * Distancetohighmountainlog10 + (1 + (Distancetosealog10 * Distancetohighmountainlog10) | Macroarea), family = "bernoulli", data = languagesonlandmasswithhighmountainsmodeling, warmup = 2000, iter = 3000, chains = 4, prior = priorshighmountains, seed = 2102, control = list(adapt_delta = 0.999999, max_treedepth = 20))
```

Model assessment and checks

check Rhat and ESS values
```{r}
summary(isolateswidehighmountainslog10model)
```
Inspect chains
```{r}
plot(isolateswidehighmountainslog10model)
```
Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(isolateswidehighmountainslog10model)
pp_check(isolateswidehighmountainslog10model, type = "error_binned")
```
Assess predictive accuracy
```{r, cache=TRUE}
isolateswidehighmountainslog10model_pred <- predict(isolateswidehighmountainslog10model, type = "response")[, "Estimate"]
isolateswidehighmountainslog10model_pred <- as.numeric(isolateswidehighmountainslog10model_pred > mean(languagesonlandmasswithhighmountainsmodeling$IsolateWide))
```
```{r, cache=TRUE}
(classtab_isolateswidehighmountainslog10model <- table(predicted = isolateswidehighmountainslog10model_pred, observed = languagesonlandmasswithhighmountainsmodeling$IsolateWide))
(acc_isolateswidehighmountainslog10model <- sum(diag(classtab_isolateswidehighmountainslog10model)) / sum(classtab_isolateswidehighmountainslog10model))
```

## Bernoulli model for isolates in a narrow sense and all mountainareas with log10 transformed predictors and random effect for macroarea
```{r, cache=TRUE, warning=FALSE, message=FALSE, results='hide'}
isolatesnarrowallmountainslog10model <- brm(IsolateNarrow ~ Distancetosealog10 * Distancetomountainlog10 + (1 + (Distancetosealog10 * Distancetomountainlog10) | Macroarea), family = "bernoulli", data = languagesonlandmasswithmountains, warmup = 2000, iter = 3000, chains = 4, prior = priorsmountains, seed = 2102, control = list(adapt_delta = 0.999999, max_treedepth = 20))
```

Model assessment and checks

check Rhat and ESS values
```{r}
summary(isolatesnarrowallmountainslog10model)
```
Inspect chains
```{r}
plot(isolatesnarrowallmountainslog10model)
```
Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(isolatesnarrowallmountainslog10model)
pp_check(isolatesnarrowallmountainslog10model, type = "error_binned")
```
Assess predictive accuracy
```{r, cache=TRUE}
isolatesnarrowallmountainslog10model_pred <- predict(isolatesnarrowallmountainslog10model, type = "response")[, "Estimate"]
isolatesnarrowallmountainslog10model_pred <- as.numeric(isolatesnarrowallmountainslog10model_pred > mean(languagesonlandmasswithmountains$IsolateNarrow))
```
```{r, cache=TRUE}
(classtab_isolatesnarrowallmountainslog10model <- table(predicted = isolatesnarrowallmountainslog10model_pred, observed = languagesonlandmasswithmountains$IsolateNarrow))
(acc_isolatesnarrowallmountainslog10model <- sum(diag(classtab_isolatesnarrowallmountainslog10model)) / sum(classtab_isolatesnarrowallmountainslog10model))
```

## Bernoulli model for isolates in a wide sense and all mountainareas with log10 transformed predictors and random effect for macroarea
```{r, cache=TRUE, warning=FALSE, message=FALSE, results='hide'}
isolatewideallmountainslog10model <- brm(IsolateWide ~ Distancetosealog10 * Distancetomountainlog10 + (1 + (Distancetosealog10 * Distancetomountainlog10) | Macroarea), family = "bernoulli", data = languagesonlandmasswithmountains, warmup = 2000, iter = 3000, chains = 4, prior = priorsmountains, seed = 2102, control = list(adapt_delta = 0.999999, max_treedepth = 20))
```
check Rhat and ESS values
```{r}
summary(isolatewideallmountainslog10model)
```
Inspect chains
```{r}
plot(isolatewideallmountainslog10model)
```
Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(isolatewideallmountainslog10model)
pp_check(isolatewideallmountainslog10model, type = "error_binned")
```
```{r, cache=TRUE}
isolatewideallmountainslog10model_pred <- predict(isolatewideallmountainslog10model, type = "response")[, "Estimate"]
isolatewideallmountainslog10model_pred <- as.numeric(isolatewideallmountainslog10model_pred > mean(languagesonlandmasswithmountains$IsolateWide))
```
```{r, cache=TRUE}
(classtab_isolatewideallmountainslog10model <- table(predicted = isolatewideallmountainslog10model_pred, observed = languagesonlandmasswithmountains$IsolateWide))
(acc_isolatewideallmountainslog10model <- sum(diag(classtab_isolatewideallmountainslog10model)) / sum(classtab_isolatewideallmountainslog10model))
```