---
title: Spatial Point Pattern Test
author: "Matthias Urban"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document: default
---
# Overview

This document describes the application of the Spatial Point Pattern Test as implemented in the R package sppt to the data analyzed in the article.

# Required packages

```{r message=FALSE}
library(tidyverse)
require(sppt)
require(maptools)
require(spatstat)
require(sf)
```

# Data preparation

Load in datasets
```{r}
languagesonlandmasswithmountains <- read.csv("../Data/languagesonlandmasswithmountains.csv", header = T, encoding = "UTF-8")
languagesonlandmasswithhighmountains <- read.csv("../Data/languagesonlandmasswithhighmountains.csv", header = T, encoding = "UTF-8")
```

For this test we first need to identify those landmasses in the OpenStreetMap polygon dataset on which one or more language is spoken. These will be used as the areal units for the test.

Load Workspace with geographical objects
```{r}
load(file = "../Data/workspace.Rdata")
```
Compute which landmasses host languages, first for the full dataset
```{r, eval=FALSE}
polygonswithlanguages <- st_intersects(languagessf, worldpolygon, sparse = FALSE)
polygonswithlanguagessummary <- as.vector(c())
for (i in 1:ncol(polygonswithlanguages)) {
  polygonswithlanguagessummary <- append(polygonswithlanguagessummary, any(polygonswithlanguages[, i]))
}
```
Add results to the landmasses, so that they are classified as either hosting mountain areas or not hosting mountain areas
```{r, cache=TRUE}
worldpolygonwithlanguages <- worldpolygon[which(polygonswithlanguagessummary == TRUE), ]
```
Convert the resulting dataset and the language dataset to SpatialPolygonsDataFrame objects
```{r, cache=TRUE}
worldpolygonswithlanguagesboundary <- as_Spatial(worldpolygonwithlanguages)
languagessp <- as_Spatial(languagessf)
```

Now the same procedure for landmasses with mountain areas
```{r, eval=FALSE}
polygonswithlanguageswithmountains <- st_intersects(languagesonlandmasswithmountainssf, mountains, sparse = FALSE)
polygonswithlanguageswithmountainssummary <- as.vector(c())
for (i in 1:ncol(polygonswithlanguageswithmountains)) {
  polygonswithlanguageswithmountainssummary <- append(polygonswithlanguageswithmountainssummary, any(polygonswithlanguageswithmountains[, i]))
}
```
Add results to the landmasses, so that they are classified as either hosting mountain areas or not hosting mountain areas
```{r, cache=TRUE}
worldpolygonwithlanguageswithmountains <- mountains[which(polygonswithlanguageswithmountainssummary == TRUE), ]
```
Convert the resulting dataset and the language dataset to SpatialPolygonsDataFrame objects
```{r, cache=TRUE}
worldpolygonswithlanguageswithmountainsboundary <- as_Spatial(worldpolygonwithlanguageswithmountains)
languagesonlandmasswithmountainssp <- as_Spatial(languagesonlandmasswithmountainssf)
```
And finally the same procedure for landmasses with mountain areas
```{r, eval=FALSE}
polygonswithlanguageswithhighmountains <- st_intersects(languagesonlandmasswithmountainssf, highmountains, sparse = FALSE)
polygonswithlanguageswithhighmountainssummary <- as.vector(c())
for (i in 1:ncol(polygonswithlanguageswithhighmountains)) {
  polygonswithlanguageswithhighmountainssummary <- append(polygonswithlanguageswithhighmountainssummary, any(polygonswithlanguageswithhighmountains[, i]))
}
```
Add results to the landmasses, so that they are classified as either hosting mountain areas or not hosting mountain areas
```{r, cache=TRUE}
worldpolygonwithlanguageswithhighmountains <- highmountains[which(polygonswithlanguageswithhighmountainssummary == TRUE), ]
```
Convert the resulting dataset and the language dataset to SpatialPolygonsDataFrame objects
```{r, cache=TRUE}
worldpolygonswithlanguageswithhighmountainsboundary <- as_Spatial(worldpolygonwithlanguageswithhighmountains)
languagesonlandmasswithhighmountainssp <- as_Spatial(languagesonlandmasswithhighmountainssf)
```

# Main analysis
```{r, cache=TRUE}
isolatesnarrowonlandmasswithhighmountainssp <- languagesonlandmasswithhighmountainssp[which(languagesonlandmasswithhighmountainssp$IsolateNarrow == T), ]
nonisolatesnarrownlandmasswithhighmountainsssp <- languagesonlandmasswithhighmountainssp[which(languagesonlandmasswithhighmountainssp$IsolateNarrow == F), ]
proportiondifferencehighmountainsisolatenarrow <- sppt_diff(isolatesnarrowonlandmasswithhighmountainssp, nonisolatesnarrownlandmasswithhighmountainsssp, worldpolygonswithlanguageswithhighmountainsboundary, test = "Fisher")
summary_sppt(proportiondifferencehighmountainsisolatenarrow)
```

# Ancillary analyses

## Full dataset, isolates in a narrow sense
```{r, cache=TRUE}
isolatesnarrowsp <- languagessp[which(languagessp$IsolateNarrow == T), ]
nonisolatesnarrowsp <- languagessp[which(languagessp$IsolateNarrow == F), ]
proportiondifferencefulldataisolatesnarrow <- sppt_diff(isolatesnarrowsp, nonisolatesnarrowsp, worldpolygonswithlanguagesboundary, test = "Fisher")
summary_sppt(proportiondifferencefulldataisolatesnarrow)
```

## Full dataset, isolates in a wide sense
```{r, cache=TRUE}
isolateswidesp <- languagessp[which(languagessp$IsolateWide == T), ]
nonisolateswidesp <- languagessp[which(languagessp$IsolateWide == F), ]
proportiondifferencefulldataisolateswide <- sppt_diff(isolateswidesp, nonisolateswidesp, worldpolygonswithlanguagesboundary, test = "Fisher")
summary_sppt(proportiondifferencefulldataisolateswide)
```

## Landmasses with mountains, isolates in a narrow sense
```{r, cache=TRUE}
isolatesnarrowonlandmasswithmountainssp <- languagesonlandmasswithmountainssp[which(languagesonlandmasswithmountainssp$IsolateNarrow == T), ]
nonisolatesnarrownlandmasswithmountainsssp <- languagesonlandmasswithmountainssp[which(languagesonlandmasswithmountainssp$IsolateNarrow == F), ]
proportiondifferencemountainsisolatenarrow <- sppt_diff(isolatesnarrowonlandmasswithmountainssp, nonisolatesnarrownlandmasswithmountainsssp, worldpolygonswithlanguageswithmountainsboundary, test = "Fisher")
summary_sppt(proportiondifferencemountainsisolatenarrow)
```

## Landmasses with mountains, isolates in a wide sense
```{r, cache=TRUE}
isolateswideonlandmasswithmountainssp <- languagesonlandmasswithmountainssp[which(languagesonlandmasswithmountainssp$IsolateWide == T), ]
nonisolateswideonlandmasswithmountainsssp <- languagesonlandmasswithmountainssp[which(languagesonlandmasswithmountainssp$IsolateWide == F), ]
proportiondifferencemountainsisolatewide <- sppt_diff(isolateswideonlandmasswithmountainssp, nonisolateswideonlandmasswithmountainsssp, worldpolygonswithlanguageswithmountainsboundary, test = "Fisher")
summary_sppt(proportiondifferencemountainsisolatewide)
```

## Landmasses with alpine mountains, isolates in a wide sense
```{r, cache=TRUE}
isolateswideonlandmasswithhighmountainssp <- languagesonlandmasswithhighmountainssp[which(languagesonlandmasswithhighmountainssp$IsolateWide == T), ]
nonisolateswidenlandmasswithhighmountainsssp <- languagesonlandmasswithhighmountainssp[which(languagesonlandmasswithhighmountainssp$IsolateWide == F), ]
proportiondifferencehighmountainsisolatewide <- sppt_diff(isolateswideonlandmasswithhighmountainssp, nonisolateswidenlandmasswithhighmountainsssp, worldpolygonswithlanguageswithhighmountainsboundary, test = "Fisher")
summary_sppt(proportiondifferencehighmountainsisolatewide)
```
