---
title: Wilcoxon-Mann-Whitney Test
author: "Matthias Urban"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document: default
---
# Overview

This document describes the procedurs for carrying out the Wilcoxon-Mann-Whitney Test reported in the article.

# Required packages

```{r message=FALSE}
library(tidyverse)
```

# Data preparation

Load in datasets
```{r}
languagesonlandmasswithmountains <- read.csv("../Data/languagesonlandmasswithmountains.csv", header = T, encoding = "UTF-8")
languagesonlandmasswithhighmountains <- read.csv("../Data/languagesonlandmasswithhighmountains.csv", header = T, encoding = "UTF-8")
```
Treat Macroareas as factors
```{r}
languagesonlandmasswithmountains$Macroarea <- as.factor(languagesonlandmasswithmountains$Macroarea)
languagesonlandmasswithhighmountains$Macroarea <- as.factor(languagesonlandmasswithhighmountains$Macroarea)
```

# Main test
```{r}
isolatesnarrowonlandmasswithhighmountains <- languagesonlandmasswithhighmountains %>% filter(IsolateNarrow == T)
nonisolatesnarrowonlandmasswithhighmountains <- languagesonlandmasswithhighmountains %>% filter(IsolateNarrow == F)
wilcox.test(isolatesnarrowonlandmasswithhighmountains$Distancetosea, nonisolatesnarrowonlandmasswithhighmountains$Distancetosea, alternative = "less")
wilcox.test(isolatesnarrowonlandmasswithhighmountains$Distancetohighmountain, nonisolatesnarrowonlandmasswithhighmountains$Distancetohighmountain, alternative = "less")
```

# Ancillary test: High Mountain areas, Isolates Wide
```{r}
isolateswideonlandmasswithhighmountains <- languagesonlandmasswithhighmountains %>% filter(IsolateWide == T)
nonisolateswideonlandmasswithhighmountains <- languagesonlandmasswithhighmountains %>% filter(IsolateWide == F)
wilcox.test(isolateswideonlandmasswithhighmountains$Distancetosea, nonisolateswideonlandmasswithhighmountains$Distancetosea, alternative = "less")
wilcox.test(isolateswideonlandmasswithhighmountains$Distancetohighmountain, nonisolateswideonlandmasswithhighmountains$Distancetohighmountain, alternative = "less")
```

# Ancillary test: All mountainareas, Isolates Narrow
```{r}
isolatesnarrowonlandmasswithmountains <- languagesonlandmasswithmountains %>% filter(IsolateNarrow == T)
nonisolatesnarrowonlandmasswithmountains <- languagesonlandmasswithmountains %>% filter(IsolateNarrow == F)
wilcox.test(isolatesnarrowonlandmasswithmountains$Distancetosea, nonisolatesnarrowonlandmasswithmountains$Distancetosea, alternative = "less")
wilcox.test(isolatesnarrowonlandmasswithmountains$Distancetomountain, nonisolatesnarrowonlandmasswithmountains$Distancetomountain, alternative = "less")
```

# Ancillary test: All  Mountain areas, Isolates Wide
```{r}
isolateswideonlandmasswithmountains <- languagesonlandmasswithmountains %>% filter(IsolateWide == T)
nonisolateswideonlandmasswithmountains <- languagesonlandmasswithmountains %>% filter(IsolateWide == F)
wilcox.test(isolateswideonlandmasswithmountains$Distancetosea, nonisolateswideonlandmasswithmountains$Distancetosea, alternative = "less")
wilcox.test(isolateswideonlandmasswithmountains$Distancetomountain, nonisolateswideonlandmasswithmountains$Distancetomountain, alternative = "less")
```